Synthesizing Quotient Lenses

### 2018-03-08 papers/QuotientLenses/main.tex
sec 4 edits

\section{QRE Lenses}
\label{QRE-lenses}
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. In this section of the paper, we introduce
\textit{QRE Lenses}, a class of quotient lenses specified using QREs.

@@

Now let $c$ and $c'$ be QREs. Given a bijection $\ell : K(c) \Leftrightarrow
K(c')$, we may define a quotient lens $q : W(c)/\eqrel{c} \Leftrightarrow
W(c')/\eqrel{c'}$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c'),
\end{equation}

\noindent In other words, the get function of the $q$ first canonizes the source
data using $\canonizer(c)$ and then applies $f$ to the result, while the
put function of the $q$ first canonizes the target data using
$\canonizer(c')$ and the applies $f^{-1}$ to the result. This is our approach in
defining QRE lenses.

One potential pitfall with QRE lenses is that the combinators that
we define on them may not preserve this normal form. For example, the
composition $q' \; ; \; q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens that has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to guarantee that every QRE
lens $q$ is equivalent to a QRE lens $q'$ which has the canonical form of
a bijection with canonizers at the edges, we need to identify a set $S$ of
bijections that will ensure that QRE lenses are closed under the action of QRE
lens combinators. Our choice is to use the class of \textit{bijective lenses},
which we briefly discuss in the next section.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_2 \circ \ell_1 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}
The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

Each bijective lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where $R$ and
$S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the source
language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$ is
$\mathcal{L}(S)$. Interestingly, with the bijective lens type system, a lens
$\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell : R'
\Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms:

@@

The other typing rules for bijective lenses are given in
Figure~\ref{fig:lensrules}.

@@

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw{lift}(\ell) \sep \kw{lquot}(q, \ell) \sep
\kw{rquot}(\ell, q) \sep q \cdot q' \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses.

@@

\subsection{Semantics of QRE Lenses}

@@

\begin{figure}[ht]
\centering
\[
\begin{array}[b]{l@{\qquad}l}
\Rule{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow id(S)}
&
\begin{array}[b]{rcl}
\llbracket \kw{lift}(\ell) \rrbracket.get &=&  \llbracket \ell \rrbracket\\
\llbracket \kw{lift}(\ell) \rrbracket.put &=& \llbracket \ell \rrbracket^{-1}
\end{array}
\end{array}
\]
\[
\begin{array}[b]{c@{\qquad}l}
\Rule{q : c'  \Leftrightarrow c'' &
\wf{c} &
K(c) = W(c')}
{\kw{lquot}(c, q): c' \circ c \Leftrightarrow c''} &
\begin{array}[b]{rcl}
\llbracket \kw{lquot}(c, q) \rrbracket.get  &=& \llbracket q
\rrbracket.get \circ \canonizer(c)\\
\llbracket \kw{lquot}(c, q) \rrbracket.put &=& \llbracket q \rrbracket.put
\end{array}
\end{array}
\]

@@

\subsection{Normal Forms of QRE Lenses}
Recall that our approach in defining QRE lenses is to have each QRE lens $q: c
\Leftrightarrow c'$ be such that
\begin{align*}
\llbracket q \rrbracket.get &= \ell \circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \ell^{-1} \circ
\canonizer(c')
\end{align*}
for some bijective lens $\ell$. In other words, each QRE lens is the same as a
bijective lens with canonizers at the edges. The following theorem,
which is the main technical contribution of this paper, confirms that this
indeed is the case:

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$ (the
full details can be found in the appendix {\sam{which I need to write}}). Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) &\equiv_{c'} y_i \text{, and}\\
q.put(y_i) &\equiv_c x_i
\end{align*}

>>>>

\section{QRE Lenses}
\label{QRE-lenses}

\dpw{I believe the purpose of the preamble was to explain the structure of the subsections and why the
material presented there appears.  I tried to rewrite this to communicate that a bit more clearly.
Please edit to improve!}
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. In this section of the paper, we introduce
\textit{QRE Lenses}, a class of quotient lenses specified using QREs.

@@

One natural way to provide a language for defining bijective quotient lenses is to
(1) define a language of bijections and then (2) add quotients by allowing canonizers
to be prepended or postpended to bijections and (3) allowing composition of such quotient
lenses via the regular operators and functional composition.  Indeed, this is the program
that we follow, with Section \ref{subsec:bijective-lenses}
defining bijections and Sections\ref{subsec:qre-lenses-syntax} and
\ref{subsec:qre-lenses-semantics}
defining the quotients and their composition operators.

However, we also have a secondary objective, which is to support synthesis of lenses: When
provided with QREs describing the source and target languages, we would like to be able to
generate quotient lenses automatically.  One way to do that is to use the QREs to generate
canonizers $c$ and $c'$ and then to synthesize bijective lenses between the kernel languages
for $c$ and $c'$.  Doing so provides us with quotient lenses defined by a particular normal
form, one in which lenses are defined by a canonizer to transform source (and target) into a normal
form and a bijection to transform between canonical forms for the source and target.
However, this normal form does not allow for sequential composition (or composition via
the regular operators) of quotients. Hence, an important technical question is whether we give
up expressiveness if we restrict ourselves to this normal form.  Fortunately, as shown in
section~\ref{subsec:qre-lens-normal-forms}, we do not.

%% Now let $c$ and $c'$ be QREs. Given a bijection $\ell : K(c) \Leftrightarrow
%% K(c')$, we may define a quotient lens $q : W(c)/\eqrel{c} \Leftrightarrow
%% W(c')/\eqrel{c'}$ as follows.
%% \begin{equation}\label{normalform}
%% q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
%% \canonizer(c'),
%% \end{equation}

%% \noindent In other words, the get function of the $q$ first canonizes the source
%% data using $\canonizer(c)$ and then applies $f$ to the result, while the
%% put function of the $q$ first canonizes the target data using
%% $\canonizer(c')$ and the applies $f^{-1}$ to the result.
%% We define QRE lenses with canonizers on the edge and bijective lenses
%% in the middle because QREs define the canonizers ($c$ and $c'$) and the kernel
%% languages $K(c)$ and $K(c')$ directly.  The kernel languages can then
%% serve as inputs to a synthesis procedure for bijections~\cite{optician}.
%% Hence, from QREs alone, we are able to generate full bijective quotient
%% lenses.

%% \dpw{Check that I'm getting this right -- I replaced the previous paragraph.}
%% One potential drawback of our approach is that it does not admit
%% composition of quotient lenses ($q' \; ; \; q$) that might interleave canonizers and bijections.
%% Our key technical result is that the lack of
%% composition does not reduce the expressiveness of the language:  Any such composition can be
%% rewritten as a single quotient lens with canonizers on the edges.

%% One potential pitfall with QRE lenses is that the combinators that we define
%% on them may not preserve this normal form.
%% For example, the composition of
%% two quotient lenses $q' \; ; \; q$ of QRE lenses $q$ and $q'$ may
%% not be equivalent to
%% any quotient lens that has the canonical normal form of a bijective lens with
%% canonizers a the edges. Consequently, in order to guarantee that every QRE
%% lens $q$ is equivalent to a QRE lens $q'$ which has the canonical form of
%% a bijection with canonizers at the edges, we need to identify a set $S$ of
%% bijections that will ensure that QRE lenses are closed under the action of QRE
%% lens combinators. Our choice is to use the class of \textit{bijective lenses},
%% which we briefly discuss in the next section.

\subsection{Bijective Lenses}
\label{subsec:bijective-lenses}

We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages constructed using the following combinators. \dpw{I said ``lens combinators''
instead of ``Boomerang lens
combinators'' because Boomerang has many other combinators... wasn't sure what the perfect thing to say was.}
%The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. The $\mathit{id}(R)$ lens applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.
\begin{align*}
\llbracket R \rrbracket &= \{(r, r) \sep r \in \mathcal{L}(R)\}\\
\llbracket const(r, s) \rrbracket &= \{(r, s)\}\\
\llbracket swap(\ell, \ell') \rrbracket &= \{(s \cdot t, t' \cdot s') \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \cdot \ell' \rrbracket &= \{(s \cdot t, s' \cdot t) \sep
(s, s') \in \llbracket \ell \rrbracket \text{ and } (t, t') \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell \sep \ell' \rrbracket &= \{(s \cdot t) \sep
(s, t) \in \llbracket \ell \rrbracket \text{ or } (s, t) \in \llbracket
\ell' \rrbracket\}\\
\llbracket \ell^* \rrbracket &= \{(s_1 \cdot \ldots \cdot s_n, t_1 \cdot \ldots
\cdot t_n) \sep (s_i, t_i) \in \llbracket \ell \rrbracket \text{ for } 1
\leq i \leq n\}\\
\llbracket \ell_2 \circ \ell_1 \rrbracket &= \{(r, t) \sep \text{there exists }s
\text{ such that } (r, s) \in \llbracket \ell_1 \rrbracket \text{ and } (s, t)
\in \llbracket \ell_2 \rrbracket\}
\end{align*}

Each bijective lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where $R$ and
$S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the source
language of $\ell$ is $\mathcal{L}(R)$ and the target language of $\ell$ is
$\mathcal{L}(S)$.  The typing rules for bijective lenses are given in
Figure~\ref{fig:lensrules}.  The final rule in the figure deserves special attention.
It states that a lens
$\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell : R'
\Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms:

@@

\noindent
This rule makes it significantly more difficult to synthesize a bijective lens from its type.

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \kw{lift}(\ell) \sep \kw{lquot}(q, \ell) \sep
\kw{rquot}(\ell, q) \sep q \cdot q' \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses.

@@

\subsection{Semantics of QRE Lenses}
\label{subsec:qre-lenses-semantics}

@@

\begin{figure}[ht]
\centering
\[
\begin{array}[b]{l@{\qquad}l}
\Rule{\ell : R \Leftrightarrow S}{\kw{lift}(\ell): id(R) \Leftrightarrow id(S)}
&
\begin{array}[b]{rcl}
\llbracket \kw{lift}(\ell) \rrbracket.get &=&  \llbracket \ell \rrbracket\\
\llbracket \kw{lift}(\ell) \rrbracket.put &=& \llbracket \ell \rrbracket^{-1}
\end{array}
\end{array}
\]
%% \[
%% \begin{array}[b]{c@{\qquad}l}
%% \Rule{q : c'  \Leftrightarrow c'' &
%% \wf{c} &
%% K(c) = W(c')}
%% {\kw{lquot}(c, q): c' \circ c \Leftrightarrow c''} &
%% \begin{array}[b]{rcl}
%% \llbracket \kw{lquot}(c, q) \rrbracket.get  &=& \llbracket q
%% \rrbracket.get \circ \canonizer(c)\\
%% \llbracket \kw{lquot}(c, q) \rrbracket.put &=& \llbracket q \rrbracket.put
%% \end{array}
%% \end{array}
%% \]

@@

\subsection{Normal Forms of QRE Lenses}
\label{subsec:qre-lens-normal-forms}
Recall that our approach in defining QRE lenses is to have each QRE lens $q: c
\Leftrightarrow c'$ be such that
\begin{align*}
\llbracket q \rrbracket.get &= \ell \circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \ell^{-1} \circ
\canonizer(c')
\end{align*}
for some bijective lens $\ell$. In other words, each QRE lens is the same as a
bijective lens with canonizers at the edges. The following theorem,
which is the main technical contribution of this paper, confirms that this
indeed is the case:

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
Sketch:
% We elide a full proof of this statement, since the proof is attained via a
% straightforward induction on the derivation $q : c \Leftrightarrow c$ (the
% full details can be found in the appendix {\sam{which I need to write}}). Since
% bijective lenses are closed under strongly unambiguous regular operators, the
% tricky case is composition.
The proof follows by induction on the derivation of $q : c \Leftrightarrow c$.
The most interesting part of the proof is the case for functional composition
as we must demonstrate that it is possible to eliminate the canonizers in the
middle of the term.

@@

\fi

\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1),\allowbreak \ldots,\allowbreak (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$. We wish $q$ to map the equivalence class of $x_i$ to the
equivalence class of $y_i$ and vice versa:
\begin{align*}
q.get(x_i) &\equiv_{c'} y_i \text{, and}\\
q.put(y_i) &\equiv_c x_i
\end{align*}
